---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales"))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
opts_chunk$set("message" = FALSE)

```

```{r rdata_tables}

```



# The Importance of the Microbiota and Individual Community Members in Colorectal Cancer

\begin{center}
\vspace{25mm}

Marc A Sze${^1}$ and Patrick D Schloss${^1}$${^\dagger}$

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI




\end{center}

Co-author e-mails:

* marcsze@med.umich.edu



\newpage
\linenumbers

## Abstract

**Background.** 

**Results.** 

**Conclusions.** 


### Keywords

microbiota; colorectal cancer; polyps; treatment; risk factor.

\newpage


## Background



\newpage

## Results

***Section 1***



***Section 2***



***Section 3***



***Section 4***




\newpage

## Discussion




## Conclusion





\newpage

## Methods

***Study Design***




***16S rRNA Gene Sequencing.***



***Sequence Processing.***




***Statistical Analysis.***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical. 


***Reproducible Methods.***



\newpage


## Declarations

### Ethics approval and consent to participate

Need to fill in.

### Consent for publication

Not applicable.

### Availability of data and material

Need to fill in.

### Competing Interests

All authors declare that they do not have any relevant competing interests to report.

### Funding

Need to fill in.

### Authors' contributions

Need to fill in.

### Acknowledgements

Need to fill in.

\newpage

## References

<div id="refs"></div>

\newpage


**Table 1: **

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}

```


\newpage


**Table 2: **

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}


```


\newpage



**Figure 1: **

**Figure 2: ** 

**Figure 3: ** 

**Figure 4: **  




\newpage


**Figure S1: ** 

**Figure S2: ** 

**Figure S3: **

\newpage
