---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: null
---

# The Microbiota and Individual Community Members in Colorectal Cancer: Is There a Common Theme?

\vspace{10mm}

\begin{center}
Marc A Sze and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}


\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))

```



**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Adenoma**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")

table_s1 <- temp_t1 %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = stringr::str_replace(study, 
                                      c("brim", "zeller", "baxter", "hale"), 
                                      c("Brim", "Zeller", "Baxter", "Hale")))

colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")

kable(table_s1, format = "markdown", align = 'c', padding = 2)
```


\newpage

**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Carcinoma**

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")

table_s2 <- temp_t2 %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = stringr::str_replace(study, 
                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))


colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")

kable(table_s2, format = "markdown", align = 'c', padding = 2)
```



\newpage


**Table S3: **
\footnotesize
```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}



```
\normalsize



\newpage



\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

**Table S2: Summary of Common OTUs for All Diagnosis Groups**

\footnotesize

```{r supplemental_table, echo=FALSE, message=FALSE, warning=FALSE}


```
\normalsize
\newpage




**Table S4: Summary of Adenoma based on Surgery Received**

```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}





```


\newpage


**Table S5: Summary of mtry ranges used for the treatment and diagnosis models**

\footnotesize

```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}




```
\normalsize






