---
title: ''
csl: mbio.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=0.75in
bibliography: null
---
\pagenumbering{gobble}
```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr"))

```


**Table S1: Comparison of odds ratios calculated using alpha diversity community metrics associated with the presence of adenomas or carcinoma relative to those in individuals with normal colons using data collected from tissue samples.**

\footnotesize

```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv")) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = round(pvalue, digits = 2),
         measure = c("OTU Richness", "Shannon Diversity", "Evenness"),
         comparison = "Adenoma",
         tissue_type = "Combined")

colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure",
                                   "Tumor", "Tissue Group")



alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = round(pvalue, digits = 2),
         measure = c("OTU Richness", "Shannon Diversity", "Evenness"),
         comparison = "Carcinoma",
         tissue_type = "Unmatched")

colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure",
                                           "Tumor", "Tissue Group")


alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv")) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = round(pvalue, digits = 2),
         measure = c("OTU Richness", "Shannon Diversity", "Evenness"),
         comparison = "Carcinoma",
         tissue_type = "Matched")

colnames(alpha_RR_crc_match_tissue) <-  c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure",
                                          "Tumor", "Tissue Group")

combined_table_S3 <- alpha_RR_adn_tissue %>% bind_rows(alpha_RR_crc_unmatch_tissue, alpha_RR_crc_match_tissue)



kable(combined_table_S3, format = "markdown", align = 'c', padding = 2)
```


\normalsize

\newpage

**Table S2: Comparison of community dissimilarity between individuals with normal colons and those with adenomas and carcinomas as calculated using Bray-Curtis distance and tested using PERMANOVA.**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")

table_s1 <- temp_t1 %>%
  mutate(fstat = round(fstat, digits = 2),
         r2 = round(r2, digits = 3),
         pvalue = round(pvalue, digits = 4),
         study = stringr::str_replace(study,
                                      c("brim", "zeller", "baxter", "hale"),
                                      c("Brim", "Zeller", "Baxter", "Hale")),
         comparison = "Adenoma",
         sample_type = "Feces") %>%
  select(study, comparison, sample_type, r2, pvalue)

colnames(table_s1) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")

temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")

table_s2 <- temp_t2 %>%
  mutate(fstat = round(fstat, digits = 2),
         r2 = round(r2, digits = 3),
         pvalue = round(pvalue, digits = 4),
         study = stringr::str_replace(study,
                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"),
                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")),
         comparison = "Carcinoma",
         sample_type = "Feces") %>%
  select(study, comparison, sample_type, r2, pvalue)


colnames(table_s2) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")

combined_table_s1 <- table_s1 %>% bind_rows(table_s2)

temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")

table_s3 <- temp_t3a %>%
  bind_rows(temp_t3b) %>%
  mutate(fstat = round(fstat, digits = 2),
         r2 = round(r2, digits = 3),
         pvalue = round(pvalue, digits = 4),
         study = c("Lu", "Flemer", "Lu (Matched)"),
         comparison = "Adenoma",
         sample_type = "Tissue") %>%
  select(study, comparison, sample_type, r2, pvalue)

colnames(table_s3) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")


temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")

table_s4 <- temp_t4a %>%
  bind_rows(temp_t4b) %>%
  mutate(fstat = round(fstat, digits = 2),
         r2 = round(r2, digits = 3),
         pvalue = round(pvalue, digits = 4),
         study = c("Sanapareddy", "Burns", "Flemer", "Chen",
                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"),
         comparison = "Carcinoma",
         sample_type = "Tissue") %>%
  select(study, comparison, sample_type, r2, pvalue)



colnames(table_s4) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")

combined_table_s2 <- table_s3 %>% bind_rows(table_s4)

combined_table_s1 <- combined_table_s1 %>% bind_rows(combined_table_s2)


kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
```

\newpage




**Table S3: ORs for individual taxa associated with individuals who had a normal colon or adenomas or carcinomas using data collected from stool.** The listed P-values were less than 0.05 prior to using a Benjimani-Hochberg correction for multiple comparisons.

\scriptsize

```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}

temp_8 <- read_csv("../data/process/tables/adn_select_genus_OR_stool_composite.csv") %>%
  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>%
  arrange(bh, pvalue, rr) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = format(pvalue, scientific = T, digits = 3),
         bh = format(bh, scientific = T, digits = 3),
         measure = str_replace_all(measure, "_unclassified", ""),
         comparison = "Adenoma") %>%
  select(measure, comparison, rr, ci_lb, ci_ub, pvalue, bh)

colnames(temp_8) <- c("Taxon", "Tumor", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")

temp_10 <- read_csv("../data/process/tables/select_genus_OR_stool_composite.csv") %>%
  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>%
  arrange(bh, pvalue, rr) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = format(pvalue, scientific = T, digits = 3),
         bh = format(bh, scientific = T, digits = 3),
         measure = str_replace_all(measure, "_unclassified", ""),
         comparison = "Carcinoma") %>%
  select(measure, comparison, rr, ci_lb, ci_ub, pvalue, bh)

colnames(temp_10) <- c("Taxon", "Tumor", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")

combined_table_S4 <- temp_8 %>% bind_rows(temp_10)

kable(combined_table_S4, format = "markdown", align = 'c', padding = 2)

```

\normalsize

\newpage


**Table S4: ORs for individual taxa associated with individuals who had a normal colon or adenomas or carcinomas using data collected from tissue samples.** The listed P-values were less than 0.05 prior to using a Benjimani-Hochberg correction for multiple comparisons.


\scriptsize

```{r supplemental_table9, echo=FALSE, message=FALSE, warning=FALSE}

temp_9 <- read_csv("../data/process/tables/adn_select_genus_OR_tissue_composite.csv") %>%
  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>%
  arrange(bh, pvalue, rr) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = format(pvalue, scientific = T, digits = 3),
         bh = format(bh, scientific = T, digits = 3),
         measure = str_replace_all(measure, "_unclassified", ""),
         comparison = "Adenoma",
         tissue_type = "Combined") %>%
  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)

colnames(temp_9) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")

temp_11 <- read_csv("../data/process/tables/select_genus_OR_unmatched_tissue_composite.csv") %>%
  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>%
  arrange(bh, pvalue, rr) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = format(pvalue, scientific = T, digits = 3),
         bh = format(bh, scientific = T, digits = 3),
         measure = str_replace_all(measure, "_unclassified", ""),
         comparison = "Carcinoma",
         tissue_type = "Unmatched") %>%
  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)

colnames(temp_11) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")

temp_12 <- read_csv("../data/process/tables/select_genus_OR_matched_tissue_composite.csv") %>%
  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>%
  arrange(bh, pvalue, rr) %>%
  mutate(rr = round(rr, digits = 2),
         ci_lb = round(ci_lb, digits = 2),
         ci_ub = round(ci_ub, digits = 2),
         pvalue = format(pvalue, scientific = T, digits = 3),
         bh = format(bh, scientific = T, digits = 3),
         measure = str_replace_all(measure, "_unclassified", ""),
         comparison = "Carcinoma",
         tissue_type = "Matched") %>%
  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)

colnames(temp_12) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")

combined_table_S5 <- temp_9 %>% bind_rows(temp_11, temp_12)

kable(combined_table_S5, format = "markdown", align = 'c', padding = 0)

```
