---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: null
---

# Investigating the Microbiota and Colorectal Cancer: The Importance of Community

\vspace{10mm}

\begin{center}
Marc A Sze and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}


\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))

```



**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")

table_s1 <- temp_t1 %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = stringr::str_replace(study, 
                                      c("brim", "zeller", "baxter", "hale"), 
                                      c("Brim", "Zeller", "Baxter", "Hale")))

colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")

kable(table_s1, format = "markdown", align = 'c', padding = 2)
```


\newpage

**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")

table_s2 <- temp_t2 %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = stringr::str_replace(study, 
                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))


colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")

kable(table_s2, format = "markdown", align = 'c', padding = 2)
```



\newpage


**Table S3: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Adenoma**

```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")

table_s3 <- temp_t3a %>% 
  bind_rows(temp_t3b) %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = c("Lu", "Flemer", "Lu (Matched)"))



colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")

kable(table_s3, format = "markdown", align = 'c', padding = 2)
```

\newpage

**Table S4: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Carcinoma**

```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")

table_s4 <- temp_t4a %>% 
  bind_rows(temp_t4b) %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"))



colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")

kable(table_s4, format = "markdown", align = 'c', padding = 2)



```











